Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2025-05-20 manuscript/manuscript.tex
Replace dG with dA
Plots not done yet

%%% Shortcuts and macros
\newcommand{\dgsolv}{\Delta G_\textup{solv}}
\newcommand{\software}{\texttt}
\newcommand{\code}{\texttt}
\newcommand{\todo}{\textcolor{red}}
\newcommand\inlinecode{\texttt}

@@

Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}. For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. 
Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

@@

\pagebreak
\section{Tutorial}
Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
The tutorial is structured as follows:
\begin{enumerate}
	\item Set up and run GIST calculations for a small molecule solute (biotin) and a protein (streptavidin) binding pocket.
	\item Compute biotin's free energy of solvation using GIST.
	\item Visualize local contributions to $\dgsolv$ in the streptavidin binding pocket and around biotin.
	\item Compute the hydration contribution of the biotin, streptavidin and complex to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
	\item Explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
\end{enumerate}
In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output. This provides an example of a simple GIST workflow, without too much details.
Afterwards, we apply this to a protein binding pocket and estimate a free energy of ligand binding using an end-states approach based on GIST. In addition, we discuss all steps shown in the first example in further detail.

@@

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is defined for rigid solute molecules.
In practice, this means that the solute molecule should not move significantly during the MD simulation.
This is typically achieved with harmonic positional restraints applied to the heavy atoms of the solute, tethering them to their initial or reference coordinates (e.g., from a crystal structure or an equilibrated structure). 
In AMBER,  the harmonic restraint potential takes the form $U = k_r * (r - r_0)^2$ for each cartesian coordinate of each restrained atom, where $r_0$ is its reference position. 
The force constant $k_r$ should be chosen strong enough to prevent significant deviation.
Here, we apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. 
Note that this is a rather strong restraint. 
While weaker restraints are possible, make sure that the solute molecule is kept rigid during the simulation. 
As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
Similarly, the solute could also be kept rigid using positional constraints, RMSD restraints, or other methods.
However, depending on the method, additional post-processing may be required to keep the solute at the same position in the grid.
To make sure that the solute is not moving significantly, we recommend to check the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
The RMSD should be less than \SI{0.5}{\angstrom} for the solute heavy atoms, preferably less than \SI{0.1}{\angstrom}.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

@@

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a \code{.loc} property similar to the pandas library. 
GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
Indexing with \code{.loc} works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}
Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.\\
Information about the grid is stored in the grid property of the gist object and structural data (supplied as a pdb of the solute molecule) is stored as an \software{mdtraj} trajectory in the struct property:
\begin{lstlisting}[style=python]
gist.grid # Grid object
gist.struct # mdtraj.Trajectory object
\end{lstlisting}
Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data. For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):

@@

\subsubsection{Dealing with Conformational Heterogeneity}
Crystal structures may not represent the dominant solution structure for a given target.
In our example, we only use a single equilibrated structure.
However, in practice the target may be flexible and a single structure may not be representative of the conformational ensemble.
A more sophisticated approach would identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
When the probabilities of the different conformations are known, one can then use a weighted average of the GIST results to obtain a free energy of solvation for the ensemble \cite{Waibl2021-gist-antibodies}:
\begin{equation}
	\Delta G_{solv} = k_BT \cdot ln \int p(q_u) \cdot e^{ \frac{-\Delta G_solv(q_u)}{k_BT}} dq_u
\end{equation}
While averaging over multiple conformations has been shown to improve agreement with experimental observables \cite{Kamenik2020-gist-macrocycles,Waibl2021-gist-antibodies}, there are some caveats to this approach. 
First, running GIST analyses for each conformation is computationally expensive and time-consuming, since each conformation needs their own MD simulation.
In cases where the conformational ensemble is large, it may be difficult to sample all relevant conformations. 
If accurate solvation thermodynamics are desired for such cases, other methods which inherently include the solute confromational ensemble may be more appropriate.
Second, this approach suffers from the curse of exponential averaging, where the free energy of solvation is dominated by very few conformations due to the exponential reweighting \cite{Ekberg2021-exponential}.
For cases where the number of sampled conformations is small and the samples are badly chosen, this can lead to large errors in the free energy of solvation.
Often a simple weighted average is then a better estimator:
\begin{equation}
	\Delta G_{solv} = \frac{\sum_i p(q_u) \cdot \Delta G_solv(q_u)}{\sum_i p(q_u)}
\end{equation}
While dealing with conformational heterogeneity is therefore definitely possible with GIST, it is not trivial and remains challenging.
However, GIST is at its best when the solute conformational ensemble is small and well-defined, such as in the case of small molecules or rigid protein-ligand complexes.
It remains unmatched when one is interested in the local solvation properties of a single dominant conformation, e.g. to elucidate the solvation thermodynamics of crystal waters in a x-ray structure.

@@

\begin{lstlisting}[style=python]
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt
gist = load_gist_file(
    'gist.dat', struct='solute-centered.pdb')
bins, eww = gist.rdf(
    'Eww_unref_norm', bins=100, rmax=20.,
    normalize='norm')
plt.plot(bins, eww)
eww_auto = gist.detect_reference_value()
plt.axhline(eww_auto, color='k')
plt.gca().set(
    xlabel='distance [A]', ylabel='E_ww per mol')
plt.show()
\end{lstlisting}
The expected output of this code is shown in Figure~\ref{fig_ewwref}.
%TODO redo figure caption, yaxis label
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin). The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
\end{figure}
After choosing an appropriate reference value, we need to subtract this value from $E_{ww}$.
Make sure to subtract the reference from the normalized (\_norm) data.
In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.

\newcommand{\coordinate}{\mathbf{r}}
\begin{equation}
	\dgsolv(\coordinate) = \Delta E_{ww}(\coordinate) + \Delta E_{sw}(\coordinate) - T\Delta S^\text{six}(\coordinate)
\end{equation}
%In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
After that, check whether your free energy contributions become negligible far away from the solute molecule.
In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence (see Figure~\ref{fig_radial_convergence}).
If the curve flattens out, the converged value is your final $\dgsolv$\@.
If the curve diverges even for large integration distances, it depends on which of the contributing parts diverges. 
\begin{lstlisting}[style=python]
	from gisttools.gist import load_gist_file
	import matplotlib.pyplot as plt
	import numpy as np
	# adapt eww_ref according to the used solvent model! 
	gist = load_gist_file('gist.dat',
		struct='solute-centered.pdb', eww_ref=-9.5398)
	bins, (dg, esw, eww, s) = gist.rdf(
		['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
		bins=100, rmax=20., normalize='none')
	plt.plot(bins, np.cumsum(eww), label='$\Delta$E$_{ww}$')
	plt.plot(bins, np.cumsum(esw), label='$\Delta$E$_{sw}$')
	plt.plot(bins, np.cumsum(dg), label='$\Delta$G')
	plt.plot(bins, np.cumsum(s), label='T$\Delta$S')
	plt.legend()
	plt.xlabel("distance cutoff [$\AA$]")
	plt.ylabel("free energy contribution [kcal/mol]")
	\end{lstlisting}
The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_noref.png}
	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule. All quantities are cumulative, i.e., summed up to the respective radius. Computed from the biotin calculation.}\label{fig_radial_convergence}
\end{figure}
Here we see two effects: First, that the solute-solvent energy $E_{sw}$ converges only slowly, due to the charged nature of biotin.
Second, that the solvent entropy diverges, even at large distances.
Most of the time, this happens due to a bias in the Nearest-Neighbor entropy calculation\cite{Chen2021,Huggins2014-entropy}.
This can be fixed by either calculating the bias per water molecule (as was done in \cite{Chen2021}) or referencing the entropy in the same manner as the $E_{ww, norm}$:
\begin{lstlisting}[style=python]
	def reference_entropy(gf):
    if 'dTSsix_unref_norm' not in gf.data.columns:
        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
    refval = gf.detect_reference_value('dTSsix_unref_dens')
    gf['dTSsix_norm'] = gf.get_referenced('dTSsix_unref_norm', refval)
    gf['dTSsix_dens'] = gf.get_referenced('dTSsix_unref_dens', refval)

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_ref.png}
	\caption{Convergence of $\dgsolv$ and its contributions with increasing distance to the solute molecule. The entropy was referenced, so it converges to zero for voxels far away from the solute.}\label{fig_radial_convergence_ref}
\end{figure}

\subsubsection{Visualizing $\dgsolv$}
Next, we visualize the enthalpic contributions to the free energy of solvation.
You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
This quantity is given by $2E_{ww} + E_{sw}$.
The energy referencing leads to non-zero normalized values where the population is zero.
While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
An OpenDX file can be produced using \software{gisttools} as follows:

@@

\subsubsection{Contribution of Hydration to Binding}
\label{sec:binding_contributions}
In the next step, calculate the $\dgsolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
Note that our integration region does not fully reach into bulk, and some of the boundary regions are close to streptavidin. 
Therefore, our results for each system will depend on the exact position of the integration boundary. 
To avoid inconsistencies, it is important to choose exactly the same integration region for each system.
Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
Therefore, we align the complex structure to streptavidin and use the shifted biotin coordinates to define the integration region.
We recommend using a \software{Jupyter Notebook} for the following analyses.
Load the files and double check the frame numbers and reference density.
\begin{lstlisting}[style=python]
import numpy as np
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt

@@

\begin{lstlisting}[style=python]
difference = compl_rdf - biotin_rdf - strept_rdf
cutoff = 12
integral = difference[bins < cutoff].sum()
print('Integral = {}'.format(integral))
plt.plot(bins, np.cumsum(difference))
plt.axvline(cutoff)
plt.xlabel('distance to biotin [Å]')
plt.ylabel('dG contribution [kcal/mol]')
\end{lstlisting}
Finally, you can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:

>>>>

%%% Shortcuts and macros
\newcommand{\dasolv}{\Delta A_\textup{solv}}
\newcommand{\software}{\texttt}
\newcommand{\code}{\texttt}
\newcommand{\todo}{\textcolor{red}}
\newcommand\inlinecode{\texttt}

@@

Methods based on IST use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}. For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches currently can only be used with sufficiently rigid solvents, although a recent paper provides the theory needed to generalize to flexible solvent molecules \cite{gilson2024freeenergy}.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. 
Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties. 
Additionally, like most other MD- and force-field based methods, a number of approximations are assumed, such as the pairwise additivity of the intermolecular interactions, the neglect of polarization and quantum effects, as well as the neglect of the solute's internal degrees of freedom.
Recent perspectives on the various limitations inherent to the force-field based MD approach were provided by \cite{van2021effect} and \cite{van2024methods}.

@@

\pagebreak
\section{Tutorial}
Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
The tutorial is structured as follows:
\begin{enumerate}
	\item Set up and run GIST calculations for a small molecule solute (biotin) and a protein (streptavidin) binding pocket.
	\item Compute biotin's free energy of solvation using GIST.
	\item Visualize local contributions to $\dasolv$ in the streptavidin binding pocket and around biotin.
	\item Compute the hydration contribution of the biotin, streptavidin and complex to the binding. Combined with the biotin--streptavidin interaction energy, the binding free energy can be estimated.
	\item Explore advantages and disadvantages of the GIST method when used to calculate $\dasolv$
\end{enumerate}
In the first part, we show how to run GIST calculations for small molecule solutes and interpret the output. This provides an example of a simple GIST workflow, without too much details.
Afterwards, we apply this to a protein binding pocket and estimate a free energy of ligand binding using an end-states approach based on GIST. In addition, we discuss all steps shown in the first example in further detail.

@@

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is defined for rigid solute molecules.
In practice, this means that the solute molecule should not move significantly during the MD simulation.
This is typically achieved with harmonic positional restraints applied to the heavy atoms of the solute, tethering them to their initial or reference coordinates (e.g., from a crystal structure or an equilibrated structure). 
In AMBER,  the harmonic restraint potential takes the form $U = k_r * (r - r_0)^2$ for each cartesian coordinate of each restrained atom, where $r_0$ is its reference position. 
The force constant $k_r$ should be chosen strong enough to prevent significant deviation.
Here, we apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. 
Note that this is a rather strong restraint. 
While weaker restraints are possible, make sure that the solute molecule is kept rigid during the simulation. 
As a minimum value, we recommend \SI{10}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
Similarly, the solute could also be kept rigid using positional constraints, RMSD restraints, or other methods.
However, depending on the method, additional post-processing may be required to keep the solute at the same position in the grid.
To make sure that the solute is not moving significantly, we recommend checking the root mean square deviation (RMSD) of the solute heavy atoms during the simulation.
The RMSD should be less than \SI{0.5}{\angstrom} for the solute heavy atoms, preferably less than \SI{0.1}{\angstrom}.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

@@

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a \code{.loc} property similar to the pandas library. 
GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
Indexing with \code{.loc} works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}
Gisttools also calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting T$\Delta$Ssix from Eall.\\
Information about the grid is stored in the grid property of the gist object and structural data (supplied as a pdb of the solute molecule) is stored as an \software{mdtraj} trajectory in the struct property:
\begin{lstlisting}[style=python]
gist.grid # Grid object
gist.struct # mdtraj.Trajectory object
\end{lstlisting}
Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data. For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):

@@

\subsubsection{Dealing with Conformational Heterogeneity}
Crystal structures may not represent the dominant solution structure for a given target.
In our example, we only use a single equilibrated structure.
However, in practice the target may be flexible and a single structure may not be representative of the conformational ensemble.
A more sophisticated approach would identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
When the probabilities of the different conformations are known, one can then use a weighted average of the GIST results to obtain a free energy of solvation for the ensemble \cite{Waibl2021-gist-antibodies}:
\begin{equation}
	\Delta A_{solv} = k_BT \cdot ln \int p(q_u) \cdot e^{ \frac{-\Delta A_{solv}(q_u)}{k_BT}} dq_u
\end{equation}
While averaging over multiple conformations has been shown to improve agreement with experimental observables \cite{Kamenik2020-gist-macrocycles,Waibl2021-gist-antibodies}, there are some caveats to this approach. 
First, running GIST analyses for each conformation is computationally expensive and time-consuming, since each conformation needs their own MD simulation.
In cases where the conformational ensemble is large, it may be difficult to sample all relevant conformations. 
If accurate solvation thermodynamics are desired for such cases, other methods which inherently include the solute confromational ensemble may be more appropriate.
Second, this approach suffers from the curse of exponential averaging, where the free energy of solvation is dominated by very few conformations due to the exponential reweighting \cite{Ekberg2021-exponential}.
For cases where the number of sampled conformations is small and the samples are badly chosen, this can lead to large errors in the free energy of solvation.
Often a simple weighted average is then a better estimator:
\begin{equation}
	\Delta A_{solv} = \frac{\sum_i p(q_u) \cdot \Delta A_{solv}(q_u)}{\sum_i p(q_u)}
\end{equation}
While dealing with conformational heterogeneity is therefore definitely possible with GIST, it is not trivial and remains challenging.
However, GIST is at its best when the solute conformational ensemble is small and well-defined, such as in the case of small molecules or rigid protein-ligand complexes.
It remains unmatched when one is interested in the local solvation properties of a single dominant conformation, e.g. to elucidate the solvation thermodynamics of crystal waters in a x-ray structure.

@@

\begin{lstlisting}[style=python]
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt
gist = load_gist_file(
    'gist.dat', struct='solute-centered.pdb')
bins, eww = gist.rdf(
    'Eww_unref_norm', bins=100, rmax=20.,
    normalize='norm')
plt.plot(bins, eww)
eww_auto = gist.detect_reference_value()
plt.axhline(eww_auto, color='k')
plt.gca().set(
    xlabel='distance [A]', ylabel='E_ww per mol')
plt.show()
\end{lstlisting}
The expected output of this code is shown in Figure~\ref{fig_ewwref}.
%TODO redo figure caption, yaxis label
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin). The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
\end{figure}
After choosing an appropriate reference value, we need to subtract this value from $E_{ww}$.
Make sure to subtract the reference from the normalized (\_norm) data.
In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
Now, sum all free energy contributions to obtain the spatially resolved $\dasolv$.
In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.

\newcommand{\coordinate}{\mathbf{r}}
\begin{equation}
	\dasolv(\coordinate) = \Delta E_{ww}(\coordinate) + \Delta E_{sw}(\coordinate) - T\Delta S^\text{six}(\coordinate)
\end{equation}
%In \software{gistpp}, use the \inlinecode{add} command, with the same syntax as the \inlinecode{div} command used above.
After that, check whether your free energy contributions become negligible far away from the solute molecule.
In a plot like Figure~\ref{fig_ewwref} based on the \inlinecode{A\_dens} column, the values should tend to zero.
However, it is more informative to plot the cumulative free energy contribution against the distance to the solute molecule, to check the convergence (see Figure~\ref{fig_radial_convergence}).
If the curve flattens out, the converged value is your final $\dasolv$\@.
If the curve diverges even for large integration distances, it depends on which of the contributing parts diverges. 
\begin{lstlisting}[style=python]
	from gisttools.gist import load_gist_file
	import matplotlib.pyplot as plt
	import numpy as np
	# adapt eww_ref according to the used solvent model! 
	gist = load_gist_file('gist.dat',
		struct='solute-centered.pdb', eww_ref=-9.5398)
	bins, (da, esw, eww, s) = gist.rdf(
		['A_dens', 'Esw_dens', 'Eww_dens', 'dTSsix_dens'],
		bins=100, rmax=20., normalize='none')
	plt.plot(bins, np.cumsum(eww), label='$\Delta$E$_{ww}$')
	plt.plot(bins, np.cumsum(esw), label='$\Delta$E$_{sw}$')
	plt.plot(bins, np.cumsum(da), label='$\Delta$A)
	plt.plot(bins, np.cumsum(s), label='T$\Delta$S')
	plt.legend()
	plt.xlabel("distance cutoff [$\AA$]")
	plt.ylabel("free energy contribution [kcal/mol]")
	\end{lstlisting}
The expected output of this code is shown in Figure~\ref{fig_radial_convergence}.
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_noref.png}
	\caption{Convergence of $\dasolv$ and its contributions with increasing distance to the solute molecule. All quantities are cumulative, i.e., summed up to the respective radius. Computed from the biotin calculation.}\label{fig_radial_convergence}
\end{figure}
Here we see two effects: First, that the solute-solvent energy $E_{sw}$ converges only slowly, due to the charged nature of biotin.
Second, that the solvent entropy diverges, even at large distances.
Most of the time, this happens due to a bias in the Nearest-Neighbor entropy calculation\cite{Chen2021,Huggins2014-entropy}.
This can be fixed by either calculating the bias per water molecule (as was done in \cite{Chen2021}) or referencing the entropy in the same manner as the $E_{ww, norm}$:
\begin{lstlisting}[style=python]
	def reference_entropy(gf):
    if 'dTSsix_unref_norm' not in gf.data.columns:
        gf['dTSsix_unref_norm'] = gf['dTSsix_norm']
        gf['dTSsix_unref_dens'] = gf['dTSsix_dens']
    refval = gf.detect_reference_value('dTSsix_unref_dens')
    gf['dTSsix_norm'] = gf.get_referenced('dTSsix_unref_norm', refval)
    gf['dTSsix_dens'] = gf.get_referenced('dTSsix_unref_dens', refval)

@@

\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/biotin-contributions_ref.png}
	\caption{Convergence of $\dasolv$ and its contributions with increasing distance to the solute molecule. The entropy was referenced, so it converges to zero for voxels far away from the solute.}\label{fig_radial_convergence_ref}
\end{figure}

\subsubsection{Visualizing $\dasolv$}
Next, we visualize the enthalpic contributions to the free energy of solvation.
You can use PyMol\cite{pymol} or VMD\cite{vmd} to visualize the .dx files from GIST\@.
With \software{gisttools}, you can create one using e.g., \inlinecode{gist.save\_dx('A\_dens', 'A\_dens.dx')}.
It might also be interesting to visualize the average energy of a water solvent molecule at each grid voxel.
This quantity is given by $2E_{ww} + E_{sw}$.
The energy referencing leads to non-zero normalized values where the population is zero.
While this is irrelevant for further post-processing, it is advantageous for visualization to set those empty regions to zero.
An OpenDX file can be produced using \software{gisttools} as follows:

@@

\subsubsection{Contribution of Hydration to Binding}
\label{sec:binding_contributions}
In the next step, calculate the $\dasolv$ contributions around biotin in each system (biotin, streptavidin, and complex).
Note that our integration region does not fully reach into bulk, and some of the boundary regions are close to streptavidin. 
Therefore, our results for each system will depend on the exact position of the integration boundary. 
To avoid inconsistencies, it is important to choose exactly the same integration region for each system.
Although we kept the systems rigid, the center of mass might have been shifted during pressure equilibration.
Therefore, we align the complex structure to streptavidin and use the shifted biotin coordinates to define the integration region.
We recommend using a \software{Jupyter Notebook} for the following analyses.
Load the files and double check the frame numbers and reference density.
\begin{lstlisting}[style=python]
import numpy as np
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt

@@

\begin{lstlisting}[style=python]
difference = compl_rdf - biotin_rdf - strept_rdf
cutoff = 12
integral = difference[bins < cutoff].sum()
print('Integral = {}'.format(integral))
plt.plot(bins, np.cumsum(difference))
plt.axvline(cutoff)
plt.xlabel('distance to biotin [Å]')
plt.ylabel('$\Delta A$ contribution [kcal/mol]')
\end{lstlisting}
Finally, you can also plot the difference between the complex and monomer contributions against the distance to biotin, using the complex coordinates for the holo structure:
